---
title: "Species Distribution Example with Random Forest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Species Distribution Example with Random Forest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10.5,
  fig.height = 6.5,
  out.width = "100%",
  fig.align = "center"
)

species_example <- requireNamespace("disdat", quietly = TRUE) &&
  requireNamespace("randomForest", quietly = TRUE)
```

This vignette shows a compact species distribution workflow using
[`disdat`](https://cran.r-project.org/package=disdat) data, a down-sampled
random forest classifier, and `curves` response plots.

The example uses one species from New South Wales (`"nsw14"`), combines its
presence records with 10,000 random background points, and then down-samples
the background class during model fitting so each tree sees a balanced sample.

The code below is shown in full, and is only evaluated when the optional
packages `disdat` and `randomForest` are installed.

## Prepare the training data

```{r, eval = species_example, message = FALSE, warning = FALSE}
spID <- "nsw14"
pr <- disdat::disPo("NSW")
bg <- disdat::disBg("NSW")

pr <- pr[pr$spid == spID, ]
training <- rbind(pr, bg)

covars <- c(
  "cti", "disturb", "mi", "rainann", "raindq",
  "rugged", "soildepth", "soilfert", "solrad",
  "tempann", "topo", "vegsys"
)

training <- training[, c("occ", covars)]
training$vegsys <- as.factor(training$vegsys)

for (v in covars[covars != "vegsys"]) {
  meanv <- mean(training[[v]])
  sdv <- stats::sd(training[[v]])
  training[[v]] <- (training[[v]] - meanv) / sdv
}

head(training[, 1:5])
```

## Fit a down-sampled random forest

`randomForest::randomForest()` can balance each bootstrap sample with
`sampsize`. Here the number of background samples drawn for each tree is set to
the number of presences, which is a common way to reduce class imbalance.

```{r, eval = species_example, message = FALSE, warning = FALSE}
training$occ <- as.factor(training$occ)
pr_num <- as.integer(table(training$occ)[["1"]])
smpsize <- c("0" = pr_num, "1" = pr_num)

set.seed(32639)
rf_model <- randomForest::randomForest(
  occ ~ .,
  data = training,
  ntree = 1000,
  sampsize = smpsize,
  replace = TRUE
)

plot(rf_model)
```

## Plot univariate and bivariate response curves

For binary classification models, `predict(..., type = "prob")` returns one
column per class. In this example the presence class is `"1"`, so the response
plots use `response = "1"` to make that choice explicit.

```{r, eval = species_example, message = FALSE, warning = FALSE}
dat <- training[, covars]

curves::univariate(
  rf_model,
  dat,
  type = "prob",
  response = "1",
  ylab = "Relative likelihood"
)
```

The next example follows the original workflow closely by plotting the third and
fourth predictors in `covars`, which correspond to `mi` and `rainann`.

```{r, eval = species_example, message = FALSE, warning = FALSE}
curves::bivariate(
  rf_model,
  dat,
  pairs = c(3, 4),
  plot_type = "heatmap",
  type = "prob",
  response = "1",
  zlab = "Relative likelihood"
)
```

For a 3D surface, a small wrapper around `predict()` is still convenient when
you want direct control over the returned probability vector.

```{r, eval = species_example && requireNamespace("plotly", quietly = TRUE), message = FALSE, warning = FALSE}
rf_prob <- function(model, newdata) {
  predict(model, newdata, type = "prob")[, "1"]
}

curves::bivariate(
  rf_model,
  dat,
  pairs = c(3, 4),
  plot_type = "surface",
  fun = rf_prob,
  zlab = "Relative likelihood"
)
```
